---
title: "Bajade_Denisse_Lab3"
author: "Denisse Gabrielle C. Bajade"
date: "2026-09-30"
output: html_document
---

# Student Information

**Name:** Denisse Gabrielle C. Bajade  
**Course:** BS Biology  
**Laboratory Exercise:** 3

# Github link: https://github.com/dcbajade/BIO118

# BIO 118 Laboratory Exercise 3

# 3A Tidying Data Using `tidyr`

## 3.2 Messy Dataset

```{r}
library(tidyr)
library(dplyr)
library(stringr)
library(forcats)
library(lubridate)

data <- data.frame(
  site = c("Site_1", "Site_1", "Site_2", "Site_2"),
  week = c(1, 2, 1, 2),
  algae_type = c(
    "Green algae",
    "Green algae",
    "Brown algae",
    "Brown algae"
  ),
  nitrate_phosphate = c(
    "3.5_0.9",
    "4.1_1.2",
    "5.3_1.5",
    "NA_NA"
  ),
  algae_density = c(50, 55, 45, NA),
  temperature = c(20, 22, NA, 25)
)

print(data)
```

## 3.3 Split and Combine Columns

### 3.3.1 `separate()`

```{r}
separated_data <- separate(
  data,
  nitrate_phosphate,
  into = c("nitrate", "phosphate"),
  sep = "_"
)

print(separated_data)
```

### 3.3.2 `unite()`

```{r}
united_data <- unite(
  separated_data,
  site_week,
  site,
  week,
  sep = "_"
)

print(united_data)
```

## 3.4 Handle Missing Values

### 3.4.1 `drop_na()`

```{r}
cleaned_data <- drop_na(
  data,
  algae_density,
  temperature
)

print(cleaned_data)
```

## 3.5 Reshape Data

### 3.5.1 `pivot_longer()`

```{r}
long_data <- pivot_longer(
  separated_data,
  cols = c(nitrate, phosphate),
  names_to = "nutrient",
  values_to = "concentration"
)

print(long_data)
```

### 3.5.2 `pivot_wider()`

```{r}
wide_data <- pivot_wider(
  long_data,
  names_from = nutrient,
  values_from = concentration
)

print(wide_data)
```

## 3.6 Streamline Data Cleaning Steps

```{r}
tidy_data <- data %>%
  separate(
    nitrate_phosphate,
    into = c("nitrate", "phosphate"),
    sep = "_"
  ) %>%
  pivot_longer(
    cols = c(nitrate, phosphate),
    names_to = "nutrient",
    values_to = "concentration"
  ) %>%
  drop_na(concentration)

print(tidy_data)
```

## 3.7 Practice: Simulated Gene Expression Dataset

```{r}
gene_expression <- data.frame(
  gene_id = c("Gene1", "Gene2", "Gene3", "Gene4"),
  ctrl_brain_rep1 = c(12, 18, 23, 9),
  ctrl_brain_rep2 = c(11, 19, 24, 10),
  treat_brain_rep1 = c(15, 21, 28, 11),
  treat_brain_rep2 = c(14, 22, 30, 12),
  ctrl_liver_rep1 = c(22, 30, 35, 20),
  ctrl_liver_rep2 = c(21, 31, 36, 21),
  treat_liver_rep1 = c(26, NA, 39, 25),
  treat_liver_rep2 = c(27, 34, 40, 26)
)

print(gene_expression)
```

### Task 1: Convert Wide Format to Long Format

```{r}
long_gene_expression <- pivot_longer(
  gene_expression,
  cols = -gene_id,
  names_to = "measurement",
  values_to = "expression"
)

print(long_gene_expression)
```

### Task 2: Split Measurement into Three Columns

```{r}
tidy_gene_expression <- separate(
  long_gene_expression,
  measurement,
  into = c("condition", "tissue", "replicate"),
  sep = "_"
)

print(tidy_gene_expression)
```

### Task 3: Identify the Missing Expression Value

```{r}
tidy_gene_expression[is.na(tidy_gene_expression$expression), ]
```

The missing expression value is for Gene2 in the Treatment condition, Liver tissue, replicate 1. It should not be removed automatically because the missing value may still be useful and can be handled as NA in later analyses.

### Task 4: Recreate `tidy_gene_expression` Using One Pipe

```{r}
tidy_gene_expression <- gene_expression %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "measurement",
    values_to = "expression"
  ) %>%
  separate(
    measurement,
    into = c("condition", "tissue", "replicate"),
    sep = "_"
  )

print(tidy_gene_expression)
```

# 3B Transforming Data Using `dplyr`

## 3.2 Dataset

```{r}
gene_data <- data.frame(
  gene_id = c(
    "Gene1", "Gene1", "Gene2",
    "Gene2", "Gene3", "Gene3"
  ),
  tissue = c(
    "Brain", "Liver", "Brain",
    "Liver", "Brain", "Liver"
  ),
  condition = c(
    "Control", "Control", "Treatment",
    "Treatment", "Control", "Treatment"
  ),
  expression = c(
    10.2, 11.5, 13.3,
    12.8, 14.9, 15.5
  ),
  replicate = c(1, 2, 1, 2, 1, 2)
)

print(gene_data)
```

## 3.3 Select and Filter Data

### 3.3.1 `select()`

```{r}
selected_data <- select(
  gene_data,
  gene_id,
  tissue,
  expression
)

print(selected_data)
```

### 3.3.2 `filter()`

```{r}
filtered_data <- filter(
  gene_data,
  tissue == "Brain",
  expression > 12
)

print(filtered_data)
```

## 3.4 Create or Modify Variables

### 3.4.1 `mutate()`

```{r}
mutated_data <- mutate(
  gene_data,
  log_expression = log(expression)
)

print(mutated_data)
```

## 3.5 Reorder Observations

### 3.5.1 `arrange()`

```{r}
arranged_data <- arrange(
  gene_data,
  expression
)

print(arranged_data)
```

## 3.6 Summarize Data

### 3.6.1 `summarize()`

```{r}
summary_data <- gene_data %>%
  group_by(tissue) %>%
  summarize(
    mean_expression = mean(expression)
  )

print(summary_data)
```

### 3.6.2 `count()`

```{r}
count_data <- count(
  gene_data,
  gene_id
)

print(count_data)
```

## 3.7 Identify Unique Observations

### 3.7.1 `distinct()`

```{r}
distinct_data <- distinct(
  gene_data,
  gene_id,
  tissue
)

print(distinct_data)
```

## 3.8 Change Column Names

### 3.8.1 `rename()`

```{r}
renamed_data <- rename(
  gene_data,
  gene_expression = expression
)

print(renamed_data)
```

## 3.9 Combine Datasets

```{r}
gene_info <- data.frame(
  gene_id = c("Gene1", "Gene2", "Gene4"),
  description = c(
    "Apoptosis-related",
    "Cell cycle control",
    "Unknown"
  )
)

print(gene_info)
```

### 3.9.1 `left_join()`

```{r}
left_join_data <- left_join(
  gene_data,
  gene_info,
  by = "gene_id"
)

print(left_join_data)
```

### 3.9.2 `full_join()`

```{r}
full_join_data <- full_join(
  gene_data,
  gene_info,
  by = "gene_id"
)

print(full_join_data)
```

## 3.10 Streamline Data Transformation

```{r}
processed_gene_data <- gene_data %>%
  filter(expression > 12) %>%
  group_by(tissue) %>%
  summarize(
    mean_expression = mean(expression),
    n = n()
  ) %>%
  arrange(desc(mean_expression))

print(processed_gene_data)
```

## 3.11 Practice: Simulated Abundance Dataset

```{r}
set.seed(123)

species_list <- c(
  "Species_A",
  "Species_B",
  "Species_C",
  "Species_D"
)

region_list <- c(
  "Region_1",
  "Region_2",
  "Region_3"
)

years <- sample(
  2000:2020,
  100,
  replace = TRUE
)

abundance <- sample(
  1:500,
  100,
  replace = TRUE
)

temperature <- sample(
  15:35,
  100,
  replace = TRUE
)

abundance_data <- data.frame(
  species = sample(
    species_list,
    100,
    replace = TRUE
  ),
  region = sample(
    region_list,
    100,
    replace = TRUE
  ),
  year = years,
  abundance = abundance,
  temperature = temperature
)

print(abundance_data)
```

### Practice Task 1: Filter Abundance > 100 and Temperature > 20

```{r}
filtered_abundance <- filter(
  abundance_data,
  abundance > 100,
  temperature > 20
)

print(filtered_abundance)
```

### Practice Task 2: Select Species, Region, and Abundance

```{r}
selected_abundance <- select(
  abundance_data,
  species,
  region,
  abundance
)

print(selected_abundance)
```

### Practice Task 3: Arrange by Species and Abundance

```{r}
arranged_abundance <- arrange(
  abundance_data,
  species,
  desc(abundance)
)

print(arranged_abundance)
```

### Practice Task 4: Calculate Overall Total Abundance and Average Temperature

```{r}
overall_summary <- summarize(
  abundance_data,
  total_abundance = sum(abundance),
  average_temperature = mean(temperature)
)

print(overall_summary)
```

### Practice Task 5: Group by Species and Calculate Mean Abundance

```{r}
species_summary <- abundance_data %>%
  group_by(species) %>%
  summarize(
    mean_abundance = mean(abundance)
  )

print(species_summary)
```

### Practice Task 6: Complete Workflow Using Pipe Operator

```{r}
regional_summary <- abundance_data %>%
  filter(year > 2010) %>%
  group_by(region) %>%
  summarize(
    mean_abundance = mean(abundance),
    mean_temperature = mean(temperature)
  ) %>%
  arrange(desc(mean_abundance))

print(regional_summary)
```

# 3C Wrangling Data Using `stringr`, `forcats`, and `lubridate`

## 3.2 Manipulate Strings Using `stringr`

```{r}
bio_data <- data.frame(
  species = c(
    "Homo sapiens ",
    "Pan troglodytes",
    " Canis lupus",
    "Homo sapiens",
    "Mus musculus"
  ),
  gene = c(
    "BRCA1-001",
    "TP53.002",
    "MYC-001",
    "BRCA2-001",
    "EGFR.003"
  ),
  condition = c(
    "Control",
    "treatment",
    "control",
    "treatment",
    "control"
  ),
  replicate = c(
    "Replicate 1",
    "Replicate 2",
    "Replicate 1",
    "Replicate 3",
    "Replicate 1"
  )
)

print(bio_data)
```

### 3.2.1 `str_detect()`

```{r}
bio_data <- bio_data %>%
  mutate(
    is_human = str_detect(species, "Homo")
  )

print(bio_data)
```

### 3.2.2 `str_replace()`

```{r}
bio_data <- bio_data %>%
  mutate(
    condition = str_replace(
      condition,
      "control",
      "Control"
    ),
    condition = str_replace(
      condition,
      "treatment",
      "Treatment"
    )
  )

print(bio_data)
```

### 3.2.3 `str_trim()`

```{r}
bio_data <- bio_data %>%
  mutate(
    species = str_trim(species)
  )

print(bio_data)
```

### 3.2.4 `str_to_lower()`

```{r}
bio_data <- bio_data %>%
  mutate(
    condition = str_to_lower(condition)
  )

print(bio_data)
```

## 3.3 Working with Factors Using `forcats`

```{r}
fish_data <- data.frame(
  species = factor(
    c(
      "Salmon", "Trout", "Bass", "Salmon",
      "Bass", "Trout", "Salmon", "Catfish"
    )
  ),
  habitat = factor(
    c(
      "River", "Lake", "River", "Lake",
      "Ocean", "River", "River", "Lake"
    )
  ),
  treatment = c(
    "Control", "Treatment", "Control", "Treatment",
    "Treatment", "Control", "Control", "Treatment"
  )
)

print(fish_data)
```


```{r}
fish_data <- fish_data %>%
  mutate(
    treatment = factor(treatment)
  )

levels(fish_data$treatment)
```

### 3.3.1 `fct_relevel()`

```{r}
levels(fish_data$treatment)

fish_data1 <- fish_data %>%
  mutate(
    treatment = fct_relevel(
      treatment,
      "Treatment",
      "Control"
    )
  )

levels(fish_data1$treatment)
```

### 3.3.2 `fct_infreq()`

```{r}
levels(fish_data$species)

fish_data7 <- fish_data %>%
  mutate(
    species = fct_infreq(species)
  )

levels(fish_data7$species)
```

## 3.4 Working with Date and Time Using `lubridate`

```{r}
bloom_data <- data.frame(
  species = c(
    "Rose", "Lily", "Tulip",
    "Sunflower", "Daisy", "Rose"
  ),
  region = c(
    "North", "South", "East",
    "West", "North", "East"
  ),
  bloom_date = c(
    "2022-04-15",
    "2022-05-10",
    "2022-04-20",
    "2022-06-01",
    "2022-04-25",
    "2022-04-30"
  ),
  bloom_time = c(
    "06:30:00",
    "08:15:00",
    "07:00:00",
    "09:45:00",
    "06:45:00",
    "07:30:00"
  )
)

print(bloom_data)
```

### 3.4.1 `ymd()`

```{r}
bloom_data <- bloom_data %>%
  mutate(
    bloom_date = ymd(bloom_date)
  )

print(bloom_data)

str(bloom_data)
```

### 3.4.2 `ymd_hms()`

```{r}
bloom_data <- bloom_data %>%
  mutate(
    bloom_datetime = ymd_hms(
      paste(bloom_date, bloom_time)
    )
  )

print(bloom_data)
```

### 3.4.3 `year()`, `month()`, `day()`, and `hour()`

```{r}
bloom_data <- bloom_data %>%
  mutate(
    year = year(bloom_datetime),
    month = month(bloom_datetime),
    day = day(bloom_datetime),
    hour = hour(bloom_datetime)
  )

print(bloom_data)
```

### 3.4.4 `difftime()`

```{r}
bloom_data <- bloom_data %>%
  mutate(
    days_since_first_bloom = as.numeric(
      difftime(
        bloom_datetime,
        min(bloom_datetime),
        units = "days"
      )
    )
  )

print(bloom_data)
```

## 3.5 Practice: Simulated Cell Biology Experiment Data

```{r}
cell_data <- data.frame(
  cell_line = c(
    "HEK293 ",
    "HeLa",
    "A549 ",
    "HepG2",
    "HeLa",
    "HEK293"
  ),
  gene_expression = c(
    "BRCA1_001",
    "TP53_002",
    "MYC_003",
    "EGFR_001",
    "BRCA2_004",
    "EGFR_005"
  ),
  treatment = c(
    "Control",
    "control",
    "Treatment",
    "TREATMENT",
    "Control",
    "control"
  ),
  observation_time = c(
    "2023-07-15 08:30:00",
    "2023-07-15 10:45:00",
    "2023-07-16 09:00:00",
    "2023-07-16 11:20:00",
    "2023-07-17 12:15:00",
    "2023-07-17 13:00:00"
  )
)

print(cell_data)
```

### Practice: Clean and Standardize String Data

```{r}
cell_data <- cell_data %>%
  mutate(
    cell_line = str_trim(cell_line),
    treatment = str_to_lower(treatment)
  )

print(cell_data)
```

### Practice: Convert and Reorder Factor Levels

```{r}
cell_data <- cell_data %>%
  mutate(
    treatment = factor(treatment),
    cell_line = factor(cell_line),
    treatment = fct_relevel(
      treatment,
      "treatment",
      "control"
    ),
    cell_line = fct_infreq(cell_line)
  )

print(cell_data)

levels(cell_data$treatment)
levels(cell_data$cell_line)
```

### Practice: Work with Date and Time Data

```{r}
cell_data <- cell_data %>%
  mutate(
    observation_time = ymd_hms(observation_time)
  )

print(cell_data)

class(cell_data$observation_time)
```

### Extract Year, Month, Day, and Hour

```{r}
cell_data <- cell_data %>%
  mutate(
    year = year(observation_time),
    month = month(observation_time),
    day = day(observation_time),
    hour = hour(observation_time)
  )

print(cell_data)
```

### Calculate Time Difference in Hours

```{r}
cell_data <- cell_data %>%
  mutate(
    hours_since_first_observation = as.numeric(
      difftime(
        observation_time,
        min(observation_time),
        units = "hours"
      )
    )
  )

print(cell_data)
```

# 3D Organizing Files within an R Project

## Check Working Directory

```{r}
getwd()
```

## Import `iris_data.csv`

```{r}
iris_data <- read.csv(
  "Lab3D_Organizing/Data/iris_data.csv"
)

head(iris_data)
```

## Get Only the `versicolor` Data

```{r}
versicolor_data <- subset(
  iris_data,
  Species == "versicolor"
)

print(versicolor_data)
```

## Save `versicolor` Data as a New CSV

```{r}
write.csv(
  versicolor_data,
  file = "Lab3D_Organizing/Data/versicolor_data.csv",
  row.names = FALSE
)
```

## Create the Plot

```{r}
plot(
  versicolor_data$Sepal.Length,
  versicolor_data$Sepal.Width,
  xlab = "Sepal Length (cm)",
  ylab = "Sepal Width (cm)",
  pch = 19
)
```

# Conclusion

In this laboratory exercise, I practiced tidying, transforming, and organizing biological data using R. I used `tidyr` for separating, combining, reshaping, and handling missing values; `dplyr` for filtering, selecting, modifying, summarizing, and joining data; `stringr` and `forcats` for cleaning strings and managing factors; and `lubridate` for working with dates and times. I also practiced organizing files within an R project by importing, filtering, saving, and plotting the iris dataset.